---
title: "Esketamine for Major Depressive Disorder with Psychotic Features"
subtitle: "Documentation for the statistical analysis plan"
author: "Breno Souza-Marques"
output: pdf_document
geometry: "left=2.5cm, right=2.5cm, top=2cm, bottom=2cm"
fontsize: 12pt
bibliography: mylib.bib
link-citations: yes
csl: apa.csl
linestretch: 1.25
header-includes:
   - \usepackage[default]{sourcesanspro}
   - \usepackage[T1]{fontenc}
mainfont: SourceSansPro
---
```{r echo=F, include = F}
knitr::opts_chunk$set(echo = TRUE, dev="cairo_pdf")
```

```{r echo=F, include=F}
library(faux)
library(tidyverse)
library(renv)
library(lmerTest)
```
# Justification

This document should be viewed as a companion to the protocol for a randomized, double-blind, placebo-controlled, parallel clinical trial of sub-anesthetic esketamine for Major Depressive Disorder with psychotic features - ESKETPD, which is published elsewhere (REFERENCE). It is primarily justified by our intent to follow the principles of transparency, openness and reproducibility proposed by the Open Science movement [@cruwellSevenEasySteps2019].

Therefore, this document has two main goals: to further explain the analysis plan and the sample size justification of the clinical trial, and second to make the underlying code openly available. This `.PDF` document was generated by a reproducible `R Markdown` script [@xieMarkdownDefinitiveGuide2022], which is built on R [@rcoreteamLanguageEnvironmentStatistical2020]. This script contains the underlying code which can be opened and inspected within R, or at the Open Science Framework repository where it is permanently registered with a digital object identifier [@souza-marquesOpenRepositoryRandomized2022].

A brief description of the study design and its relation to our modelling strategy for the statistical analysis plan is discussed. Use of simulation and the assumptions of our analysis plan for this study are presented in detail, followed by a step-by-step process of how we conducted such simulations, and finally a brief discussion of their implications to the study.

# Statistical analysis plan

## Study design and aim

The study aims to detect possible differences in depressive symptoms between patients treated with intravenous esketamine and those treated with midazolam as an active placebo, in a trial with two parallel arms. Further information on the study methods, including participants and instruments, are available on the protocol (REFERENCE). The main outcome for this analysis is the mean Montgomery-Asberg Depression Rating Scale (MADRS) score at 24h after the last esketamine infusion. The MADRS is a 10-item scale that ranges from 0 to 60, which is reliably used in clinical trials of antidepressants, including rapid-acting antidepressants such as ketamine [@khanRelativeSensitivityMontgomeryAsberg2002; @coyleUseKetamineAntidepressant2015].

## Modelling strategy

To attain the study objectives, we focus on the difference in depressive symptoms between periods (baseline vs post-treatment) and also between groups (esketamine vs placebo). We are thus focused on the change in MADRS scores for those treated with esketamine in comparison to those treated with midazolam, which by iteself has no known antidepressant effects and acts as an active placebo comparison. This means that the chosen model should contain the factors of Period, Drug, and their interaction, which is the main result of this study.

"Drug" is considered a main factor that varies "between-participants", that is, between groups. Period is the other main factor, which varies "within-participants", from study start to end. These are the main predictos for the dependent variable, which by itself varies within each study participant from baseline to post-treatment. More specifically, this clinical trial focuses on the possible effects between levels of Drug (that is, Esketamine vs Placebo) and levels  of Period (that is, 24h post-treatment vs baseline). A model of choice needs to accommodate factors that vary between and within-participants - hierarchy, and a dependent variable collected twice from the same participant - dependency between measures.

Given these characteristics and the available statistical tools, a Linear Mixed-Effects Model (LMM) is chosen as a modelling strategy for its ability to accommodate both the dependency and hierarchical structure of the study design - for further information on LMMs, see [@mageziLinearMixedeffectsModels2015; @bolkerGeneralizedLinearMixed2009]. The main model for this study features MADRS scores as the dependent variable; Period, Drug and their interaction as fixed effects, and Participants as random effects, as follows:

## Using simulations to determine sample size

The sample size for this study was established by means of simulation, which allowed precise determination of the required number of participants given a set of predetermined parameters. 

# Study rationale







```{r cars}
summary(cars)
```

\newpage

# References