---
title: "Esketamine for Major Depressive Disorder with Psychotic Features"
subtitle: "Documentation for the statistical analysis plan"
author:
 - Breno Souza-Marques
 - Ana Teresa Caliman-Fontes
 - Gustavo C. Leal
 - Rodrigo P. Mello
 - Flávia Vieira
 - Aline S. Sampaio
 - Lucas C. Quarantini^[Corresponding Author]
output: 
               pdf_document:
                              latex_engine: xelatex
                              toc: TRUE
                              toc_depth: 3
mainfont: Times New Roman
geometry: "left=2.5cm, right=2.5cm, top=2cm, bottom=2cm"
fontsize: 12pt
bibliography: mylib.bib
link-citations: yes
csl: apa.csl
linestretch: 1.25

---
```{r echo=F, include = F}
knitr::opts_chunk$set(echo = TRUE, dev="cairo_pdf")
```

```{r echo=F, include=F}
library(faux)
library(tidyverse)
library(renv)
library(lmerTest)
```
\newpage
## Justification

This document should be viewed as a companion to a protocol for a randomized, double-blind, placebo-controlled, parallel clinical trial of sub-anesthetic esketamine for Major Depressive Disorder with psychotic features - ESKETPD, which is published elsewhere (REFERENCE). It is primarily justified by our intent to follow the principles of transparency, openness and reproducibility proposed by the Open Science movement [@cruwellSevenEasySteps2019; @kathawallaEasingOpenScience2021]. Therefore, this document has two main goals: to further explain the analysis plan and sample size of the clinical trial, and to make the underlying code openly available. This `.PDF` document was generated by a reproducible `R Markdown` script [@xieMarkdownDefinitiveGuide2022], which is built on R [@rcoreteamLanguageEnvironmentStatistical2020]. This script contains the underlying code which can be opened and inspected within R, or at the Open Science Framework repository where it is permanently registered with a digital object identifier [@souza-marquesOpenRepositoryRandomized2022].

A brief description of the study design and its relation to our modelling strategy for the statistical analysis plan is presented. Use of simulation and the assumptions of our analysis plan are presented in detail, followed by a step-by-step process of how we conducted such simulations, and finally a brief discussion of their implications to the study design.

## Study description

The study aims to detect possible differences in depressive symptoms between patients treated with intravenous esketamine and those treated with midazolam as an active placebo, in a trial with two parallel arms. Further information on the study methods, including participants and instruments, are available on the protocol (REFERENCE). The main outcome for this analysis is the mean Montgomery-Asberg Depression Rating Scale (MADRS) score at 24h after the last esketamine infusion. The MADRS is a 10-item scale that ranges from 0 to 60, which is reliably used in clinical trials of antidepressants, including rapid-acting antidepressants such as ketamine [@khanRelativeSensitivityMontgomeryAsberg2002; @coyleUseKetamineAntidepressant2015].

## Modelling strategy

We focus on the difference in depressive symptoms between periods (baseline vs post-treatment) and groups (esketamine vs placebo), that is, the change in MADRS scores for those treated with esketamine in comparison to those treated with midazolam. This means that the chosen model should contain the factors of Period, Drug, and their interaction. The "Drug" factor varies between participants, allocated in one of two groups, while "Period" varies within participants, in line with the repeated measuring of depressive symptoms as per study design.

Given these characteristics and the available statistical tools, a Linear Mixed-Effects Model (LMM) is chosen as a modelling strategy for its ability to accommodate both the dependency (data points at baseline and post-treatment are not independent) and hierarchical (multilevel) structure of the study design - for further information on LMMs, see [@mageziLinearMixedeffectsModels2015; @bolkerGeneralizedLinearMixed2009]. More specifically, we chose a LMM with a random intercept for participants, as limitations on sample size would not allow for a model with random intercepts and slopes. For the simulations of power and sample size, we assume the dependent variable, MADRS scores, will follow a somewhat symmetric shape, and thus the LMM is run with a Gaussian distribution and an identity link.

The main model for this study features MADRS scores as the dependent variable; Period, Drug and their interaction as fixed effects, and Participants as random effects. A simplified LMM model is as follows: $y = X\beta + Z\mu + e$. In this example, $X\beta$ encodes the fixed effects of Period and Drug, $Z\mu$ the random effects, and $e$ the error term. In R code, this model is classically written as `MADRS ~ Period * Drug + (1 | Subj_ID)` [@batesFittingLinearMixedEffects2015]. Further exploratory analyses on the effects of covariates of interest, such as age and sex, will follow the same structure but are not a part of the sample size calculations.

## Statistical power for LLMs through simulations

The need to determine specific parameters that exert influence on the 

The sample size for this study was established by means of simulation, which allowed precise determination of the required number of participants given a set of predetermined parameters.







```{r cars}
summary(cars)
```

\newpage

# References